---
title: "Svalbard Midtre Lovenbreen -- Zhou modeling study -- July 2021"
subtitle: "Biogeochemistry - TOC, TN, TP, TS, NO3, PO4, SO4, and microbial cell count"
author: "Adam J. Solon"
date: "`r Sys.Date()`"
#output: html_document
output: 
  pdf_document:
    toc: TRUE
    fig_width: 7
    fig_height: 6
    fig_caption: true
fontsize: 12pt
#editor_options: 
#  chunk_output_type: console
---

# Script Summary  
This script summarizes the biogeochemical measurements of samples from soils collected for the SUN SPEARS project.  The overarching question: 

What are the mean concentrations at the four sample locations?

### Steps of this pipeline:  
1.  Create and organize directories
2.  Load R packages
3.  Input files
4.  Format Files
5.  Summary statistics 
6. Save files and figures
  
```{r echo = FALSE, include = FALSE, set.seed(461)}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(include = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(cache = TRUE)

```

# Begin pipeline

### Set paths for Taxonomic Summary analyses  
```{r, echo = FALSE, include = FALSE}
# Change identifiers to your system and file naming. 
user <- "~" # computer user name
folder <- "/Desktop"
project <- "/Svalbard/studies"
study <- "/Chronosequence"
files <- "/data"

directory.1 <- "/analyses/amplicons/16s"
sub.directory.1.1 <- "/input"
sub.directory.1.2 <- "/taxa_summary"

ecosystem <- "glacial forefield - Svalbard" # Define the environment for use in report language.

```

```{r set paths for pipeline}
# First define the project and project folders. 

# Create pathway for pipeline
###################################################
path.fp <- paste0(user, folder, project, study, files)
if (!dir.exists(path.fp)) dir.create(path.fp)

```

```{r set path for analyses}
#directory.1
###################################################
directory.1.fp <- paste0(path.fp, directory.1)
if (!dir.exists(directory.1.fp)) dir.create(directory.1.fp)

# Create sub-directory  sub.directory.1.1
###################################################
sub.directory.1.1.fp <- paste0(directory.1.fp, sub.directory.1.1)
if (!dir.exists(sub.directory.1.1.fp)) dir.create(sub.directory.1.1.fp)

# Create sub-directory  sub.directory.1.2
###################################################
sub.directory.1.2.fp <- paste0(directory.1.fp, sub.directory.1.2)
if (!dir.exists(sub.directory.1.2.fp)) dir.create(sub.directory.1.2.fp)

# Create sub-directories objects and output
###################################################

# Create directory for 'within' pipeline R objects 
objects.1.2.fp <- paste0(sub.directory.1.2.fp, "/objects")
if (!dir.exists(objects.1.2.fp)) dir.create(objects.1.2.fp, recursive = TRUE)

# Create directory for 'after' pipeline outputs 
output.1.2.fp <- paste0(sub.directory.1.2.fp, "/output")
if (!dir.exists(output.1.2.fp)) dir.create(output.1.2.fp, recursive = TRUE)

# Check for the folders here as a sanity check. Should see "objects" & "output'.
list.files(sub.directory.1.2.fp) 

```

### Session Information  
```{r Install and load packages}
# install.packages("tidyverse")
# install.packages("knitr")
# install.packages("phyloseq")

library(tidyverse); packageVersion("tidyverse")
library(knitr); packageVersion("knitr")
library(phyloseq); packageVersion("phyloseq")

```  

* r version: `r getRversion()`
* RStudio version: `r rstudioapi::versionInfo()$version`
* r packages:  
  tidyverse, `r packageVersion("tidyverse")`  
  knitr, `r packageVersion("knitr")`  
  phyloseq, `r packageVersion("phyloseq")` 
  
### Input files 
Phyloseq objects from script 'chrono-normalization-transformation-16s.Rmd' 

```{r import phyloseq object}
#import R object- rarefaction sequence counts
phy.r.2 <- readRDS("~/Desktop/Svalbard/studies/Chronosequence/data/analyses/amplicons/16s/input/phy.r.2.rds")

```

##Rarefaction (i.e., repeated random subsampling without replacement for each sample) 
```{r data transform percentage rarefy}
#transform rarefaction 'counts' into relative abundance percentage
phy.r.3 <- transform_sample_counts(phy.r.2, function(x) x / sum(x) )

#filter very low percentage
#GPfr = filter_taxa(GPr, function(x) mean(x) > 1e-5, TRUE)

```

####Plot Phyla as rarefaction data
```{r barplot Rarefaction phylum}
#plot using ggplot
#convert phyloseq object into data.frame
df.rf <- psmelt(phy.r.3)

#convert phyloseq object into metadata table - data.frame
m <- as(sample_data(phy.r.3), "data.frame")

#samples to retain
values_to_retain <- c("C11.1", "C11.2", "C11.3", "C13.1", "C13.2", "C13.3", "C15.1", "C15.2", "C15.3", "C17.1", "C17.2", "C17.3")

#filter to retain only the rows with the specified samples
m.1 <- m %>%
  filter(samples %in% values_to_retain)

# Drop unused factor levels
m.1$soil.age.f <- droplevels(m.1$soil.age.f)

#convert phyloseq object into metadata table - data.frame
m.2 <- m.1 %>% 
  dplyr::select(1, 11:12, 19, 25:26, 29, 55:57, 62)

```

```{r convert moles to grams}
#converte to micrograms
m.3 <- m.2%>%
  mutate(
    ug.C_g.soil = umol.C_g.soil * 12.011,
    ug.N_g.soil = umol.N_g.soil * 14.007,
    ug.S_g.soil = umol.S_g.soil * 32.066
  )

#remove columns with umol
m.4 <- m.3 %>%
  dplyr::select(1:4, 12, 13, 11, 14, 8:10)

```

```{r summary stats}
#define function for standard error
standard_error <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(sum(!is.na(x)))
}

#compute mean, standard error, and sample size for each variable at each sample location
biogeochem.sum.stats <- m.4 %>%
  group_by(soil.age.f) %>%
  summarise(across(where(is.numeric), list(mean = ~ round(mean(.x, na.rm = TRUE), 2), 
                                           se = ~ round(standard_error(.x), 2), 
                                           n = ~ sum(!is.na(.x))), 
                   .names = "{col}_{fn}"))

```

```{r tidy dataframe}
#select and rename specific columns to be simpler to read
biogeochem.sum.stats.2 <- biogeochem.sum.stats %>%
  select(soil.age.f,
         cell_count_per.g.soil_mean, cell_count_per.g.soil_se, cell_count_per.g.soil_n,
         ug.C_g.soil_mean, ug.C_g.soil_se, ug.C_g.soil_n,
         ug.N_g.soil_mean, ug.N_g.soil_se, ug.N_g.soil_n,
         P_aqua.regia_mean, P_aqua.regia_se, P_aqua.regia_n, 
         ug.S_g.soil_mean, ug.S_g.soil_se, ug.S_g.soil_n,
         NO3_mean, NO3_se, NO3_n, 
         PO4_mean, PO4_se, PO4_n,
         SO4_mean, SO4_se, SO4_n) %>%
  rename(Cell.count.mean = cell_count_per.g.soil_mean,
         Cell.count.se = cell_count_per.g.soil_se,
         Cell.count.n = cell_count_per.g.soil_n,
         TOC.mean = ug.C_g.soil_mean,
         TOC.se = ug.C_g.soil_se,
         TOC.n = ug.C_g.soil_n,
         TN.mean = ug.N_g.soil_mean,
         TN.se = ug.N_g.soil_se,
         TN.n = ug.N_g.soil_n,
         TP.mean = P_aqua.regia_mean,
         TP.se = P_aqua.regia_se,
         TP.n = P_aqua.regia_n,
         TS.mean = ug.S_g.soil_mean,
         TS.se = ug.S_g.soil_se,
         TS.n = ug.S_g.soil_n)

#use scientific notation for large numbers
biogeochem.sum.stats.3 <- biogeochem.sum.stats.2 %>%
  mutate(across(where(is.numeric), ~ ifelse(. > 1000, formatC(., format = "e", digits = 2), .)))

#transpose so rows are columns and columns are rows
biogeochem.sum.stats.4 <- t(biogeochem.sum.stats.3)

```

```{r save files, include = FALSE}
#check and set working directory
setwd("~/Desktop/Svalbard/collaboration_manuscripts/zhou/")
getwd()

#save raw data used for analysis
write.csv(m.4, file = "Zhou-et-al-biogeochem-data.csv", row.names = FALSE)

#save summary statistics
write.csv(biogeochem.sum.stats.3, file = "Zhou-et-al-biogeochem.sum.stats.csv", row.names = FALSE)

#save summary statistics
write.csv(biogeochem.sum.stats.4, file = "Zhou-et-al-biogeochem.sum.stats.II.csv", row.names = TRUE)

```